Making Sequential Programs Deterministic

### 2019-03-27 design.tex
Edits in section 2.1

A \textit{type} serves as an abstraction for the set of acceptable values for any expression.
A \textit{type qualifier} on a \textit{type} adds additional constraints on the values represented by that \textit{type}.
A compile time type checker can verify the correctness of this \textit{type} with its dataflow analysis and subtyping rules.
In Java, a type qualifier is written as an annotation. The type \<Integer> represents positive and negative integers besides zero.
A program analysis developer could design a type system that differentiates positive integers from the negative ones with type qualifiers
\<@Positive> and \<@Negative>.  The type \<@Positive Integer> now represents only positive values and the dataflow analysis ensures that
any negative integer value flowing into this type is reported as an error.

\begin{figure}
    $\infer[\sc{\mathit{SUBTYPING}}]{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$

@@

    $\infer[\sc{\mathit{METHOD \ CALLS}}]{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$

@@

\subsection{Determinism type hierarchy}\label{type-hierarchy}
The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

In this type system, \<Det> is a subtype of \<NonDet> denoted as \codeid{Det $\sqsubseteq$ NonDet}.

\begin{figure}
    \begin{center}
        \includegraphics[scale=0.5]{determinism}
    \end{center}
    \caption{Determinism type hierarchy}
    \label{fig-determinism-hierarchy}
\end{figure}

@@

%\todo{It is very important to express the whole section as formal type
%  jugdments, or we will get laughed
%  out of the room as amateurs, during the program committee meeting.  Also,
%  the formalism will help make the ideas precise.
%
%  Type well-formedness rules:\todo{format nicely in \LaTeX}
%
%  antecedent: $\Gamma : \kappa_e <: \kappa_c$
%
%  consequent: $\Gamma : \kappa_c \ \codeid{List}\langle\kappa_e\ 
%  \codeid{Object}\rangle$
%  }
\begin{figure}
    $\infer[\sc{\mathit{ANY\ TYPE}}]{\vdash : \kappa \  \beta}{\kappa : \mathit{Det / NonDet}}$

@@

    $\infer[\sc{\mathit{ORDERNONDET \ COLLECTION}}]{\vdash : \kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Iterator}}$

@@

    $\infer[\sc{\mathit{NONDET \ COLLECTION}}]{\vdash : \mathit{NonDet} \ \beta_c\langle\mathit{NonDet} \ \beta_e\rangle}{\beta_c \sqsubseteq \mathit{Collection/Iterator}}$

@@

     $\infer[\sc{\mathit{ORDERNONDET \ ARRAY}}]{\vdash : \kappa_e \ \beta_e \ \  \kappa_a []}{\kappa_a : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_a }$

@@

    $\infer[\sc{\mathit{NONDET \ ARRAY}}]{\vdash : \mathit{NonDet} \ \beta_e \ \  \mathit{NonDet} []}{}$

@@

    $\infer[\sc{\mathit{ORDERNONDET \ MAP}}]{\vdash : \kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}{\kappa_m: \mathit{OrderNonDet/Det} & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \mathit{Map}}$

@@

    $\infer[\sc{\mathit{NONDET \ MAP}}]{\vdash : \mathit{NonDet} \ \beta_m\langle\mathit{NonDet} \ \beta_k, \mathit{NonDet} \ \beta_v\rangle}{\beta_m \sqsubseteq \mathit{Map}}$

@@

Figure~\ref{type-validity} shows the type validity rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
where $\kappa$ is a type qualifier from the determinism type hierarchy and $\beta$ is a Java type.
As stated in the rule $\sc{\mathit{ANY\ TYPE}}$, the types \<Det> and \<NonDet> can be written on any Java type.
The (determinism) type of a Collection or Iterator must be a supertype or equal to
the type of the type parameter (Rule $\sc{\mathit{ORDERNONDET \ COLLECTION}}$) (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<NonDet>, then the type parameter may not be
\<Det> or \<OrderNonDet> (Rule $\sc{\mathit{NONDET \ COLLECTION}}$ in Figure~\ref{type-validity}). Although such types could exist, they are
disallowed to prevent the following bug:

@@

Similarly, the (determinism) type of an array must be a supertype or equal to
the type of the component type (see Figure~\ref{fig-determinism-collections}).
As with collections, \codeid{NonDet} arrays of \codeid{Det} or \codeid{OrderNonDet}
elements are not allowed (Rules $\sc{\mathit{ORDERNONDET \ ARRAY}}$ and $\sc{\mathit{NONDET \ ARRAY}}$ in Figure~\ref{type-validity}).

@@

% Not true: We allow Poly(up) and down at every location where Poly is allowed.
%Valid locations for \codeid{@PolyDet("up")} and  \codeid{@PolyDet("down")}:
%$\infer{\vdash : \kappa \  \beta}{isParam(\beta) / isReturn(\beta) & \kappa : @PolyDet("up") / @PolyDet("down")}$
%
%Resolution of \<@PolyDet("up")>:\todo{typesetting}
%
%$\infer{\Gamma \vdash x : @NonDet \ \beta_a}{\Gamma \vdash x : \kappa_a \ \beta_a &  \kappa_a : @OrderNonDet/@NonDet & \kappa_p : @PolyDet("up") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$
%
%Resolution of \<@PolyDet("down")>:
%
%$\infer{\Gamma \vdash x : @Det \  \beta_a}{\Gamma \vdash x :  \kappa_a \beta_a & \kappa_a : @OrderNonDet/@Det & \kappa_p : @PolyDet("down") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$

@@

\begin{figure}
    $\infer[\sc{\mathit{POLY RESOLUTION}}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\mathit{NonDet, OrderNonDet, Det\}} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\mathit{PolyDet}, \mathit{PolyDet}\uparrow, \mathit{PolyDet}\downarrow    \}}$
    \caption{Polymorphic resolution rule}
    \label{fig-poly-resolution}
\end{figure}

@@

The determinism checker has additional type refinement rules for the sorting methods as shown in Figure~\ref{fig-sorting}. 
\begin{figure}
    $\infer[\sc{\mathit{LIST \  SORT}}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{x.sort()}}$

@@

    $\infer[\sc{\mathit{COLLECTIONS \  SORT}}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{Collections.sort.sort(x)}}$

@@

    $\infer[\sc{\mathit{ARRAYS \  SORT}}]{\Gamma \vdash x : \mathit{Det \ \beta \ Det[]} }{\Gamma \vdash x : \mathit{Det \ \beta \ OrderNonDet[]} & \mathit{Arrays.sort( \ x)}}$

@@

\begin{figure}
    $\infer[\sc{\mathit{HASHMAP \ CREATION}}]{\mathit{new \ \kappa\ HashMap()}}{\kappa = \mathit{OrderDet/NonDet}}$

@@

    $\infer[\sc{\mathit{TREEMAP \ CREATION}}]{\mathit{new \ \kappa\ TreeMap()}}{\kappa = \mathit{Det/NonDet}}$

@@

    $\infer[\sc{\mathit{HASHSET \ CREATION}}]{\mathit{new \ \kappa\ HashSet()}}{\kappa = \mathit{OrderDet/NonDet}}$

@@

    $\infer[\sc{\mathit{TREESET \ CREATION}}]{\mathit{new \ \kappa\ TreeSet()}}{\kappa = \mathit{Det/NonDet}}$
    \caption{\<Map> and \<Set> creation rules}
    \label{fig-creation-rules}
\end{figure}

@@

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet$\uparrow$ boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{linebreak}
    $\infer[\sc{\mathit{SET \ PRECISION}}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{x} \ \beta_{x}\rangle, \mathit{y} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{y} \ \beta_{y}\rangle & \mathit{z = x.equals(y)} & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle }$

@@

    $\infer[\sc{\mathit{MAP \ PRECISION}}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x : OrderNonDet \ Map\langle Det \ \beta_k, \kappa_v\ \beta_v \rangle, y : Det \ \beta} & \mathit{z = x.get(y)}}$

>>>>

In a type system,
a \textit{type} stands for a set of values.  It abstracts or restricts the
set of possbile run-time values that an expression may evaluate to.
A programming language provides \emph{basetypes}, such as \<int>.
A \textit{type qualifier} on a \textit{type} adds additional constraints;
that is, it reduces the size of the set of values.
A type checker can verify the correctness of this \textit{type} with its dataflow analysis and subtyping rules.
In Java, a type qualifier is written as a type annotation with a leading \<@>, as in
\<@Positive int>.

\begin{figure}
    \bigskip

    $\infer{\tau_1 \sqsubseteq \tau_3}{\tau_1 \sqsubseteq \tau_2 \quad \tau_2 \sqsubseteq \tau_3}
    \quad\quad
    \infer[\textsc{subtyping}]{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$

@@

    $\infer{\kappa_1 \beta \sqsubseteq \kappa_2 \beta}{\kappa_1 \sqsubseteq \kappa_2}
    \quad\quad
    \infer{\kappa \beta_1 \sqsubseteq \kappa \beta_2}{\beta_1 \sqsubseteq \beta_2}$

    $\infer{\Det \sqsubseteq \OrderNonDet}{}
    \quad\quad
    \infer{\OrderNonDet \sqsubseteq \NonDet}{}$

    $\infer[\textsc{method call}]{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$

@@

\subsection{Determinism type hierarchy}\label{type-hierarchy}

\begin{wrapfigure}{R}{0.3\textwidth}
    \begin{center}
        \includegraphics[scale=0.5]{determinism}
    \end{center}
    \caption{Determinism type qualifier hierarchy}
    \label{fig-determinism-hierarchy}
\todo{Drop the at-signs (@).  The text is weirdly small for the box.
  Probably just redraw this.}
\end{wrapfigure}

The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

@@

\begin{figure}
    $\infer[\textsc{any type}]{\wellformed{\kappa \  \beta}}{\kappa : \mathit{Det / NonDet}}$

@@

    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Iterator}}$

@@

    $\infer[\textsc{nondet collection}]{\wellformed{\mathit{NonDet} \ \beta_c\langle\mathit{NonDet} \ \beta_e\rangle}}{\beta_c \sqsubseteq \mathit{Collection/Iterator}}$

@@

     $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_a }$

@@

    $\infer[\textsc{nondet array}]{\wellformed{\mathit{NonDet} \ \beta_e \ \  \mathit{NonDet} []}}{}$

@@

    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}}{\kappa_m: \mathit{OrderNonDet/Det} & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \mathit{Map}}$

@@

    $\infer[\textsc{nondet map}]{\wellformed{\mathit{NonDet} \ \beta_m\langle\mathit{NonDet} \ \beta_k, \mathit{NonDet} \ \beta_v\rangle}}{\beta_m \sqsubseteq \mathit{Map}}$

@@

Figure~\ref{type-validity} shows the type validity rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
where $\kappa$ is a type qualifier from the determinism type hierarchy and $\beta$ is a Java type.
As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet> can be written on any Java type.
The (determinism) type of a Collection or Iterator must be a supertype or equal to
the type of the type parameter (Rule $\textsc{ordernondet collection}$) (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<NonDet>, then the type parameter may not be
\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in Figure~\ref{type-validity}). Although such types could exist, they are
disallowed to prevent the following bug:

@@

Similarly, the (determinism) type of an array must be a supertype or equal to
the type of the component type (see Figure~\ref{fig-determinism-collections}).
As with collections, \codeid{NonDet} arrays of \codeid{Det} or \codeid{OrderNonDet}
elements are not allowed (Rules $\textsc{ordernondet array}$ and $\textsc{nondet array}$ in Figure~\ref{type-validity}).

@@

% Not true: We allow Poly(up) and down at every location where Poly is allowed.
%Valid locations for \codeid{@PolyDet("up")} and  \codeid{@PolyDet("down")}:
%$\infer{\wellformed{\kappa \  \beta}}{isParam(\beta) / isReturn(\beta) & \kappa : @PolyDet("up") / @PolyDet("down")}$
%
%Resolution of \<@PolyDet("up")>:\todo{typesetting}
%
%$\infer{\Gamma \vdash x : @NonDet \ \beta_a}{\Gamma \vdash x : \kappa_a \ \beta_a &  \kappa_a : @OrderNonDet/@NonDet & \kappa_p : @PolyDet("up") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$
%
%Resolution of \<@PolyDet("down")>:
%
%$\infer{\Gamma \vdash x : @Det \  \beta_a}{\Gamma \vdash x :  \kappa_a \beta_a & \kappa_a : @OrderNonDet/@Det & \kappa_p : @PolyDet("down") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$

@@

\begin{figure}
    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\mathit{NonDet, OrderNonDet, Det\}} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\mathit{PolyDet}, \mathit{PolyDet}\uparrow, \mathit{PolyDet}\downarrow    \}}$
    \caption{Polymorphic resolution rule}
    \label{fig-poly-resolution}
\end{figure}

@@

The determinism checker has additional type refinement rules for the sorting methods as shown in Figure~\ref{fig-sorting}. 
\begin{figure}
    $\infer[\textsc{list sort}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{x.sort()}}$

@@

    $\infer[\textsc{collections sort}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{Collections.sort.sort(x)}}$

@@

    $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \mathit{Det \ \beta \ Det[]} }{\Gamma \vdash x : \mathit{Det \ \beta \ OrderNonDet[]} & \mathit{Arrays.sort( \ x)}}$

@@

\begin{figure}
    $\infer[\textsc{hashmap creation}]{\mathit{new \ \kappa\ HashMap()}}{\kappa = \mathit{OrderDet/NonDet}}$

@@

    $\infer[\textsc{treemap creation}]{\mathit{new \ \kappa\ TreeMap()}}{\kappa = \mathit{Det/NonDet}}$

@@

    $\infer[\textsc{hashset creation}]{\mathit{new \ \kappa\ HashSet()}}{\kappa = \mathit{OrderDet/NonDet}}$

@@

    $\infer[\textsc{treeset creation}]{\mathit{new \ \kappa\ TreeSet()}}{\kappa = \mathit{Det/NonDet}}$
    \caption{\<Map> and \<Set> creation rules}
    \label{fig-creation-rules}
\end{figure}

@@

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet$\uparrow$ boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{linebreak}
    $\infer[\textsc{set precision}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{x} \ \beta_{x}\rangle, \mathit{y} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{y} \ \beta_{y}\rangle & \mathit{z = x.equals(y)} & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle }$

@@

    $\infer[\textsc{map precision}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x : OrderNonDet \ Map\langle Det \ \beta_k, \kappa_v\ \beta_v \rangle, y : Det \ \beta} & \mathit{z = x.get(y)}}$
